Sperm do not evolve to collaborate in female meiotic drive.

### 2014-05-12 writeup_and_notes/male_dependent_drive.tex
edited text

Although the allelic identity of sperm could plausibly influence the outcome of female meiosis, 
	limited gene expression in sperm \citep[e.g.][]{Joseph2004}
	suggest a model where sperm influence female meiosis via expression of the fertilizing male's
	diploid genotype (perhaps due to proteins and RNAs packaged into the sperm), rather than sperm haplotype.
%GRAHAM I HAVE REWRITTEN THIS SECTION TO REFLECT MATHS AND MY PARAMETERIZATION. WE CAN WORK ON SYNCHING OUR NOTATION
%This adds an additional parameter to the system -- the dominance of the drive enhancer in heterozygous males, 
%	when this dominance is high, a narrow sliver of parameter space sustaining a polymorphism emerges (see Figure S\ref{Effect_of_dominance})
 %presumably because positive frequency dependence diminishes with the dominance of the male effect driver.
This male-genotype dependent model requires one additional parameter, as we exchange $d$ in the sperm dependent case for $d_{het}$ and $d_{hom}$ which describe the transmission of the drive allele in a heterozygous female mating with males heterozygous and homozygous for the self-promoting drive allele, respectively.  
However, when
the distorting effect of genotype displays strong dominance in its
effect on female meiosis ($d_{het}$ is close to $d_{hom}$), 
there is a a narrow sliver of parameter space sustains a polymorphism
when the cost of the drive is recessive
(see Figure S\ref{Effect_of_dominance}). 
Presumably this zone of polymorphism comes about because positive frequency dependence diminishes with the dominance of the male effect driver. 
As an analytical example, consider the case of male-genotype dependent deleterious recessive female driver 
	(assuming, as above, $w_{BB}=1-s$, and $w_{AA}=w_{AB}=1$). 
Here, a rare driver invades when $s$ is less than $(d_{het}-1/2)/d_{het}$, and fixes when $s$ is less than $d_{hom} - 1/2$.
Intermediate values of $s$, which represent a small proportion of parameter space can maintain a polymorphism. 
However, it does not seem particularly realistic to think that the
effect of the drive allele would be dominant in its action through the
male genotype, while the cost would be recessive. Thus we regard this
case as interesting but not very general.

%\gc{Add short paragraph about sex chromosomes here? XY systems don't care. ZW systems (males ZZ, females ZW). If there is
%sex ratio distortion how could this play out?}

@@

Above, we explored a model were a driver that drove in females when signaled by a complementary allele in sperm.  
We complement this single locus model of an allele with pleiotropic effects in both sexes, 
 with  an alternative model of two linked loci - one female driver and the other a sperm allele which modifies the effect of drive upon fertilization. 
%Here, we examine an alternative scenario (and more realistic) in which
In this model, a female meiotic driver with no sperm-dependency, initially reaches drive-viability equilibrium (with two alleles
$A$ and $B$ are the ancestral non-driver and driver alleles).
Subsequently, a sperm-acting modifier of female meiotic drive arises tightly
        linked to this balanced locus (effectively creating a third
        allele/haplotype at this locus).
Assuming tight linkage between female driver and sperm modifier is consistent with the nature of well characterized meiotic drive systems 
	which are often maintained as polymorphic inversions with numerous linked modifiers (CITE).

@@

%%%sperm-acting facilitator
We first consider a sperm-acting facilitator of
	female meiotic drive tightly linked to the $B$ background, 
	generating a third allele $B^{+}$ (i.e. the non-recombining haplotype of a
        driver allele and sperm-acting drive modifier). 
This new allele acts in sperm to increase the effectiveness of both
	the $B$ and  $B^{+}$ drive in $AB$ and $AB^{+}$ heterozygotes, see Figure \ref{Eggsperm_3_allele_cartoon} 
	for a simple schematic of this drive model.  
Naively, the $B^{+}$ may spread by capitalizing on the additional drive it causes; however,  
	this is not the case for a few simple reasons. 
First, the sperm-acting drive facilitator (the novel $B^{+}$ haplotype) 
	arises when the ancestral driver ($B$) is at drive-selection balance, 
	and therefore immediately suffers a homozygous fitness cost.  
Worse yet, a novel $B^{+}$ haplotype most often helps 
	the $B$  allele drive ($B^+$ sperm meeting $AB$ eggs), because $B$ is initially more common than $B^{+}$. 
As $B^{+}$ facilitates $B$ these interactions frequently form 
	$BB^{+}$ heterozygotes that suffer the same fitness costs as $BB$. 
Therefore, sperm-acting drive facilitator alleles experience a profound disadvantage 
	in this scenario, even more so than under the previous two allele model. 
We have found no parameter range of this
	three allele system that allows the sperm-acting drive facilitator $B^{+}$ to
	invade the population. \yb{MATHS?}

@@

Why is it that an allele that biases female meiosis in its favor can generate a genetic conflict, but an allele in sperm that assists this female driver cannot? 
So long as the transmission advantage of female meiotic drive outweigh the organismal fitness cost to heterozygotes the female driver can spread when rare, and it increases in 		
	frequency until the fitness cost experienced when homozygous balances the transmission advantage.
By contrast a sperm promotor of female drive is only effective when matched with a heterozygote female -- meaning that when rare this allele rarely enhances female drive. 
Even worse, when it does so it will preferentially find itself in a low fitness drive homozygote. 
Not only are drive-promoting sperm alleles unable to create a sustained genetic conflict, 
	but alleles in sperm with the opposite effect - that is those that prevent their own drive through female meiosis do maintain a polymorphisms and 
	provide evolution  with time and opportunity to further minimize drive.
This is because such drive suppressing alleles reduced their chances
of forming low fitness homozygotes. 
In fact, an allele in sperm which causes the alternate allele to drive can be favored (FIG), a counterintuitive result 
	consistent with the tenuous evidence that the non-driving
        $ln^-$ allele is over transmitted through female meiosis in $ln^-/ln^+$ heterozygote when fertilized by $ln^+$
        sperm (CITE cite{agulnick, Pomiankowski1993}).   
More generally, alleles that act through sperm to
 reduce the opportunity of female meiotic drive regardless of linkage or phase.

@@

\gc{Is there some nice way to work in green beard ideas}

>>>>

Although the allelic identity of sperm could plausibly influence the outcome of female meiosis, 
	limited gene expression in sperm \citep[e.g.][]{Joseph2004}
	suggest a model where sperm influence female meiosis via expression of the fertilizing male's
	diploid genotype (perhaps due to proteins and RNAs packaged into the sperm), rather than sperm haplotype.
This male-genotype dependent model requires one additional parameter, as we exchange $d$ in the sperm dependent case for $d_{het}$ and $d_{hom}$ which describe the transmission of the drive allele in a heterozygous female mating with males heterozygous and homozygous for the self-promoting drive allele, respectively.  
However, when
the distorting effect of genotype displays strong dominance in its
effect on female meiosis ($d_{het}$ is close to $d_{hom}$), 
there is a a narrow sliver of parameter space sustains a polymorphism
when the cost of the drive is recessive
(see Figure S\ref{Effect_of_dominance}). 
Presumably this zone of polymorphism comes about because positive frequency dependence diminishes with the dominance of the male effect driver. 
As an analytical example, consider the case of male-genotype dependent deleterious recessive female driver 
	(assuming, as above, $w_{BB}=1-s$, and $w_{AA}=w_{AB}=1$). 
Here, a rare driver invades when $s$ is less than $(d_{het}-1/2)/d_{het}$, and fixes when $s$ is less than $d_{hom} - 1/2$.
Intermediate values of $s$, which represent a small proportion of parameter space can maintain a polymorphism. 
However, it does not seem particularly realistic to think that the
effect of the drive allele would be dominant in its action through the
male genotype, while the cost would be recessive. Thus we regard this
case as interesting but not very general.

@@

Above, we explored a model were a driver that drove in females when signaled by a complementary allele in sperm.  
We complement this single locus model of an allele with pleiotropic effects in both sexes, 
 with  an alternative model of two linked loci - one female driver and the other a sperm allele which modifies the effect of drive upon fertilization. 
In this model, a female meiotic driver with no sperm-dependency, initially reaches drive-viability equilibrium (with two alleles
$A$ and $B$ are the ancestral non-driver and driver alleles).
Subsequently, a sperm-acting modifier of female meiotic drive arises tightly
        linked to this balanced locus (effectively creating a third
        allele/haplotype at this locus). Tight linkage offers the best
        chance for a collaboration to evolve between a driver and
       an enhancer of the drive, as recombination breaks up drive haplotypes \citep{Thomson1974,Charlesworth1978,Haig1991}. 
In addition tight linkage between female driver and sperm modifier is consistent with the nature of well characterized drive systems which are often maintained as polymorphic inversions with numerous linked modifiers (CITE).

@@

%%%sperm-acting facilitator
We first consider a sperm-acting facilitator of
	female meiotic drive tightly linked to the $B$ background, 
	generating a third allele $B^{+}$ (i.e. the non-recombining haplotype of a
        driver allele and sperm-acting drive modifier). 
This new allele acts in sperm to increase the effectiveness of both
	the $B$ and  $B^{+}$ drive in $AB$ and $AB^{+}$ heterozygotes, see Figure \ref{Eggsperm_3_allele_cartoon} 
	for a simple schematic of this drive model.  
Naively, the $B^{+}$ may spread by capitalizing on the additional drive it causes; however,  
	this is not the case for a few simple reasons. 
First, the sperm-acting drive facilitator (the novel $B^{+}$ haplotype) 
	arises when the ancestral driver ($B$) is at drive-selection balance, 
	and therefore immediately suffers a homozygous fitness cost.  
Worse yet, a novel $B^{+}$ haplotype most often helps 
	the $B$  allele drive ($B^+$ sperm meeting $AB$ eggs), because $B$ is initially more common than $B^{+}$. 
As $B^{+}$ facilitates $B$ these interactions frequently form 
	$BB^{+}$ heterozygotes that suffer the same fitness costs as $BB$. 
Therefore, sperm-acting drive facilitator alleles experience a profound disadvantage 
	in this scenario, even more so than under the previous two allele model. 
We have found no parameter range of this
	three allele system that allows the sperm-acting drive facilitator $B^{+}$ to
	invade the population.

@@

Why is it that an allele that biases female meiosis in its favor can generate a genetic conflict, but an allele in sperm that assists this female driver cannot? 
So long as the transmission advantage of female meiotic drive outweigh the organismal fitness cost to heterozygotes the female driver can spread when rare, and it increases in 		
	frequency until the fitness cost experienced when homozygous balances the transmission advantage.
By contrast a sperm promotor of female drive is only effective when matched with a heterozygote female -- meaning that when rare this allele rarely enhances female drive. 
Even worse, when it does so it will preferentially find itself in a low fitness drive homozygote. 
Not only are drive-promoting sperm alleles unable to create a sustained genetic conflict, 
	but alleles in sperm with the opposite effect - that is those that prevent their own drive through female meiosis do maintain a polymorphisms and 
	provide evolution  with time and opportunity to further minimize drive.
This is because such drive suppressing alleles reduced their chances
of forming low fitness homozygotes. 
In fact, an allele in sperm which causes the alternate allele to drive can be favored (FIG), a counterintuitive result 
	consistent with the tenuous evidence that the non-driving
        $ln^-$ allele is over transmitted through female meiosis in $ln^-/ln^+$ heterozygote when fertilized by $ln^+$
        sperm \cite{agulnick, Pomiankowski1993}.   
More generally, alleles that act through sperm to
 reduce the opportunity of female meiotic drive regardless of linkage or phase.

@@
